Using R as a GIS
========================================================

Working with spatial data and basic GIS operations in R
-------------------------------------------------------
v 2.0  
15/02/2013   
Francisco Rodriguez-Sanchez  

Look for the latest version at
http://sites.google.com/site/rodriguezsanchezf


```{r}
setwd("~/UsingR-GIS")    # set working directory
```


BASIC PACKAGES
--------------

```{r message=FALSE}

library(sp)         # classes for spatial data
library(raster)     # grids, rasters
library(rasterVis)  # raster visualisation
library(maptools)   
    # and their dependencies
```

   
     
     
VISUALISING GEOGRAPHICAL DATA
=============================

  
  

RWORLDMAP: for mapping global data
----------------------------------

```{r message=FALSE, warning=FALSE}

library(rworldmap)

# examples:
newmap <- getMap(resolution="coarse")    # different resolutions available
plot(newmap)
```

```{r message=FALSE} 
mapCountryData()
```
```{r message=FALSE}
mapCountryData(mapRegion="europe")
```
```{r message=FALSE}
mapGriddedData()
```
```{r message=FALSE}
mapGriddedData(mapRegion="europe")
  
```



  
    
GOOGLEVIS: visualise data in a web browser using Google Visualisation API
-------------------------------------------------------------------------

```{r message=FALSE}

library(googleVis) 

# demo(googleVis)   # run this demo to see all the possibilities
```

```{r setOptions, echo=FALSE}
op <- options(gvis.plot.tag = "chart")  
# necessary so that googleVis works with knitr, see http://lamages.blogspot.co.uk/2012/10/googlevis-032-is-released-better.html
```


Example: plot country-level data
```{r results='asis', tidy=FALSE, eval=FALSE}
data(Exports)
View(Exports)       # a simple data frame
Geo <- gvisGeoMap(Exports, locationvar="Country", numvar="Profit", 
                  options=list(height=400, dataMode='regions'))
# print(Geo)  # this prints HTML code that can be embedded in a web page!
plot(Geo)

```


Example: Plotting point data onto a google map (internet)
```{r results='asis', tidy=FALSE, eval=FALSE}
data(Andrew)
M1 <- gvisMap(Andrew, "LatLong" , "Tip", options=list(showTip=TRUE, showLine=F, enableScrollWheel=TRUE,
                         mapType='satellite', useMapTypeControl=TRUE, width=800,height=400))
plot(M1)

```


  
    


RGOOGLEMAPS: Map your data onto Google Map tiles
------------------------------------------------

```{r message=FALSE, results='hide'}

library(RgoogleMaps)   

# First, get maps from Google (a file will be saved in your working directory)
newmap <- GetMap(center=c(36.7,-5.9), zoom =10, destfile = "newmap.png", maptype = "satellite")   

# Now using bounding box instead of center coordinates:
newmap2 <- GetMap.bbox(lonR=c(-5, -6), latR=c(36, 37), destfile = "newmap2.png", maptype="terrain")   

# Try different maptypes
newmap3 <- GetMap.bbox(lonR=c(-5, -6), latR=c(36, 37), destfile = "newmap3.png", maptype="satellite")
```

```{r}
# Now plot data onto these maps, e.g. these 3 points
PlotOnStaticMap(lat = c(36.3, 35.8, 36.4), lon = c(-5.5, -5.6, -5.8), zoom= 10, cex=2, pch= 19, col="red", FUN = points, add=F)

```



  
   
LIBRARY "DISMO": FUNCTION GMAP
------------------------------


```{r message=FALSE}

library(dismo)
  
# Some examples

# Getting maps for countries
mymap <- gmap("France")   # choose whatever country
plot(mymap)
mymap <- gmap("France", type="satellite")   # choose map type
plot(mymap)
mymap <- gmap("France", type="satellite", exp=3)  # choose the zoom level
plot(mymap)

# Save the map as a file in your wd for future use
mymap <- gmap("France", type="satellite", filename="France.gmap")    
```

```{r eval=FALSE}
# Now get a map for a region drawn at hand
mymap <- gmap("Europe")
plot(mymap)
select.area <- drawExtent()   # now click 2 times on the map to select your region
mymap <- gmap(select.area)
plot(mymap)
# See ?gmap for many other possibilities


```




SPATIAL VECTOR DATA (POINTS, POLYGONS, ETC)
===========================================




Let's create an example dataset:
```{r message=FALSE}
# Get known locations of the laurel tree (Laurus nobilis) from GBIF
library(dismo)      # check also the nice "rgbif" library! 
laurus <- gbif("Laurus", "nobilis")      
# get data frame with spatial coordinates (points)
locs <- subset(laurus, select=c("country", "lat", "lon"))
head(locs)    # a simple data frame with coordinates

# Discard data with errors in coordinates:
locs <- subset(locs, locs$lat<90)
```

Making your data 'spatial'
```{r}
coordinates(locs) <- c("lon", "lat")    # set spatial coordinates
plot(locs)
```


Define geographical projection:
```{r}
# to look for the appropriate PROJ.4 description look here:
# http://www.spatialreference.org/
crs.geo <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")    # geographical, datum WGS84
proj4string(locs) <- crs.geo     # define projection system of our data
summary(locs)
```

Plotting point data on a map
```{r}
plot(locs, pch=20, col="steelblue")
library(rworldmap)
# library rworldmap provides different types of global maps, e.g:
data(coastsCoarse) 
data(countriesLow)
plot(coastsCoarse, add=T)
```

Subsetting
```{r}
table(locs@data$country)     # see localities of Laurus nobilis by country
locs.gr <- subset(locs, locs$country=="GR")   # select only locs in Greece
plot(locs.gr, pch=20, cex=2, col="steelblue")
title("Laurus nobilis occurrences in Greece")
plot(countriesLow, add=T)
summary(locs.gr)

locs.gb <- subset(locs, locs$country=="GB")    # locs in UK
plot(locs.gb, pch=20, cex=2, col="steelblue")
title("Laurus nobilis occurrences in UK")
plot(countriesLow, add=T)

```




MAKING MAPS WITH VECTORIAL SPATIAL DATA (POINTS, POLYGONS, POLYLINES)
=====================================================================


USING RGOOGLEMAPS
-----------------

```{r message=FALSE}

require(RgoogleMaps)
locs.gb.coords <- as.data.frame(coordinates(locs.gb))    # retrieves coordinates (1st column for longitude, 2nd column for latitude)
PlotOnStaticMap(lat = locs.gb.coords$lat, lon = locs.gb.coords$lon, zoom= 5, cex=1.4, pch= 19, col="red", FUN = points, add=F)

## Downloading map from Google Maps and plotting onto it
map.lim <- qbbox (locs.gb.coords$lat, locs.gb.coords$lon, TYPE="all")    # define region of interest (bounding box)
mymap <- GetMap.bbox(map.lim$lonR, map.lim$latR, destfile = "gmap.png", maptype="satellite")
# see the file in the wd
PlotOnStaticMap(mymap, lat = locs.gb.coords$lat, lon = locs.gb.coords$lon, zoom= NULL, cex=1.3, pch= 19, col="red", FUN = points, add=F)

## using different background
mymap <- GetMap.bbox(map.lim$lonR, map.lim$latR, destfile = "gmap.png", maptype="hybrid")
PlotOnStaticMap(mymap, lat = locs.gb.coords$lat, lon = locs.gb.coords$lon, zoom= NULL, cex=1.3, pch= 19, col="red", FUN = points, add=F)

```



USING "DISMO": FUNCTION GMAP
------------------------------

```{r}
gbmap <- gmap(locs.gb, type="satellite")
locs.gb.merc <- Mercator(locs.gb)    # Google Maps are in Mercator projection. This function projects the points to that projection to enable mapping
plot(gbmap)
points(locs.gb.merc, pch=20, col="red")
```



PLOTTING ONTO A GOOGLE MAP USING GOOGLEVIS (internet)
-----------------------------------------------------

```{r results='asis', tidy=FALSE, eval=FALSE}
points.gb <- as.data.frame(locs.gb)
points.gb$latlon <- paste(points.gb$lat, points.gb$lon, sep=":")
map.gb <- gvisMap(points.gb, locationvar="latlon", tipvar="country", 
                  options = list(showTip=T, showLine=F, enableScrollWheel=TRUE,
                           useMapTypeControl=T, width=1400,height=800))
plot(map.gb)
#print(map.gb)    # get HTML suitable for a web page
```



DRAWING POLYGONS AND POLYLINES
------------------------------

```{r eval=FALSE}
plot(gbmap)
mypolygon <- drawPoly()    # click on the map to draw a polygon and press ESC when finished
summary(mypolygon)    # now you have a spatial polygon! Easy, isn't it?
```




CONVERTING, READING AND SAVING SPATIAL VECTOR DATA
==================================================


Exporting KML (Google Earth)
----------------------------
```{r}
writeOGR(locs.gb, dsn="locsgb.kml", layer="locs.gb", driver="KML")
```

Reading kml
-----------
```{r}
newmap <- readOGR("locsgb.kml", layer="locs.gb")
```

Saving as a Shapefile
---------------------
```{r}
writePointsShape(locs.gb, "locsgb")
```

Reading shapefiles
--------------------------
```{r}
gb.shape <- readShapePoints("locsgb.shp")
plot(gb.shape)

# readShapePoly         # read polygon shapefiles   
# readShapeLines        # read polylines   
# see also shapefile in "raster"   
```





CHANGING PROJECTION OF SPATIAL VECTOR DATA
==========================================

```{r}
# Projecting data frame with Laurus nobilis coordinates
summary(locs)    # original coordinates in lat lon
# define new projection; look up parameters at http://spatialreference.org
crs.laea <- CRS("+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") # Lambert Azimuthal Equal Area
locs.laea <- spTransform(locs, crs.laea)    # spTransform makes the projection

# Projecting shapefile of countries
plot(countriesLow)    # countries map in geographical projection
country.laea <- spTransform(countriesLow, crs.laea)  # project

# Plotting
plot(locs.laea, pch=20, col="steelblue")
plot(country.laea, add=T)
# define spatial limits for plotting
plot(locs.laea, pch=20, col="steelblue", xlim=c(1800000, 3900000), ylim=c(1000000, 3000000))
plot(country.laea, add=T)
```




OVERLAY
-------

Get number of occurrences (points) by country
```{r}
ov <- overlay(locs.laea, country.laea)
countr <- country.laea@data$NAME[ov]
summary(countr)




```






USING RASTER (GRID) DATA
========================



DOWNLOADING RASTER CLIMATE DATA FROM INTERNET
---------------------------------------------

```{r}
tmin <- getData("worldclim", var="tmin", res=10)   # this will download global data on minimum temperature at 10' resolution
# with this function you can also get other climatic data, elevation, administrative boundaries, etc
```


LOADING A RASTER LAYER
----------------------

```{r}
tmin1 <- raster("~/UsingR-GIS/wc10/tmin1.bil")   # Tmin for January
fromDisk(tmin1)  # values are stored on disk instead of memory! (useful for large rasters)
tmin1 <- tmin1/10    # Worldclim temperature data come in decimal degrees 
tmin1    # look at the info
plot(tmin1)

# ?raster    # raster reads many different formats, including Arc ASCII grids or netcdf files
```


CREATING A RASTER STACK
-----------------------

A raster stack is collection of many raster layers with the same projection, spatial extent and resolution

```{r message=FALSE, warning=FALSE}

library(gtools)
list.ras <- mixedsort(list.files("~/UsingR-GIS/wc10/", full.names=T, pattern=".bil"))
list.ras   # I have just collected a list of the files containing monthly temperature values
tmin.all <- stack(list.ras)
tmin.all
tmin.all <- tmin.all/10
plot(tmin.all)
```


RASTER BRICKS
-------------

A rasterbrick is similar to a raster stack (i.e. multiple layers with the same extent and resolution),
but all the data must be stored in a single file

```{r}
tmin.brick <- brick(tmin.all)   # creates rasterbrick
```



CROP RASTERS
------------

```{r eval=FALSE}

## Crop raster manually:
plot(tmin1)
newext <- drawExtent()    # click twice on the map to select the region of interest
tmin1.c <- crop(tmin1, newext)
plot(tmin1.c)
```

```{r}
## Alternatively, provide coordinates for the limits of the region of interest
newext <- c(-10, 10, 30, 50)   
tmin1.c <- crop(tmin1, newext)
plot(tmin1.c)

tmin.all.c <- crop(tmin.all, newext)
plot(tmin.all.c)
```


DEFINE SPATIAL PROJECTION OF THE RASTERS
----------------------------------------

```{r}
crs.geo    # defined above
projection(tmin1.c) <- crs.geo
projection(tmin.all.c) <- crs.geo
tmin1.c    # notice info at coord.ref.


### CHANGING PROJECTION

tmin1.proj <- projectRaster(tmin1.c, crs="+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +a=6378137 +b=6378137 +units=m +no_defs")
tmin1.proj   # notice info at coord.ref.
plot(tmin1.proj)
# can also use a template raster, see ?projectRaster
```



PLOTTING RASTER DATA
--------------------

Different plotting functions:
```{r}
histogram(tmin1.c)
pairs(tmin.all.c)
persp(tmin1.c)
contour(tmin1.c)
contourplot(tmin1.c)
levelplot(tmin1.c)
plot3D(tmin1.c)
bwplot(tmin.all.c)
densityplot(tmin1.c)
```



Spatial autocorrelation
-----------------------

```{r}
Moran(tmin1.c)    # global Moran's I
tmin1.Moran <- MoranLocal(tmin1.c)
plot(tmin1.Moran)
```


EXTRACT VALUES FROM RASTER
--------------------------

```{r}
head(locs)    # we'll obtain tmin values for our points
projection(tmin1) <- crs.geo
locs$tmin1 <- extract(tmin1, locs)    # raster values are incorporated to the dataframe
head(locs)
```

```{r eval=FALSE}
# extract values for a given region
plot(tmin1.c)
reg.clim <- extract(tmin1.c, drawExtent())  # click twice to draw extent of the region of interest 
summary(reg.clim)
```

```{r}
# rasterToPoints
tminvals <- rasterToPoints(tmin1.c)
head(tminvals)
```

```{r eval=FALSE}
## CLICK function: get values from particular locations in the map
plot(tmin1.c)
click(tmin1.c, n=3)   # click n times in the map 
```



RASTERIZE POINTS, LINES OR POLYGONS
-----------------------------------

```{r}
locs2ras <- rasterize(locs.gb, tmin1)
locs2ras
plot(locs2ras, xlim=c(-10,10), ylim=c(45, 60), legend=F)
data(wrld_simpl)
plot(wrld_simpl, add=T)
```


CHANGING RASTER RESOLUTION (aggregate)
--------------------------------------

```{r}
tmin1.lowres <- aggregate(tmin1.c, fact=2, fun=mean)
tmin1.lowres
tmin1.c     # compare
par(mfcol=c(1,2))
plot(tmin1.c, main="original")
plot(tmin1.lowres, main="low resolution")
```


SPLINE INTERPOLATION
--------------------

```{r message=FALSE, warning=FALSE}
xy <- data.frame(xyFromCell(tmin1.lowres, 1:ncell(tmin1.lowres)))    # get raster cell coordinates
head(xy)
vals <- getValues(tmin1.lowres)
library(fields)
spline <- Tps(xy, vals)    # thin plate spline
intras <- interpolate(tmin1.c, spline)
intras    # note new resolution
plot(intras)  
intras <- mask(intras, tmin1.c)   # mask to land areas only
plot(intras)
title("Interpolated raster")
```


SETTING ALL RASTERS TO THE SAME EXTENT, PROJECTION AND RESOLUTION ALL IN ONE
----------------------------------------------------------------------------

```{r message=FALSE, eval=F}
library(climstats)
spatial_sync_raster
```



ELEVATIONS, SLOPE, ASPECT, ETC
------------------------------

Download elevation data from internet:
```{r}
elevation <- getData('alt', country='ESP')
x <- terrain(elevation, opt=c('slope', 'aspect'), unit='degrees')
plot(x)

slope <- terrain(elevation, opt='slope')
aspect <- terrain(elevation, opt='aspect')
hill <- hillShade(slope, aspect, 40, 270)
plot(hill, col=grey(0:100/100), legend=FALSE, main='Spain')
plot(elevation, col=rainbow(25, alpha=0.35), add=TRUE)
```



SAVING AND EXPORTING RASTER DATA
--------------------------------

Saving raster to file:
----------------------
```{r}
writeRaster(tmin1.c, filename="tmin1.c.grd")   # can export to many different file types, see help
writeRaster(tmin.all.c, filename="tmin.all.grd")
```

Exporting to KML (Google Earth)
-------------------------------
```{r}
tmin1.c <- raster(tmin.all.c, 1)
KML(tmin1.c, file="tmin1.kml")  
KML(tmin.all.c)     # can export multiple layers


```



SPATIAL STATISTICS
==================


Point pattern analysis
----------------------


Useful packages:

```{r message=FALSE}
library(spatial)
library(spatstat)
library(spatgraphs)
library(ecespa)    # ecological focus
# etc (see Spatial Task View: http://cran.r-project.org/web/views/Spatial.html)

# Example: Ripley's K
data(fig1)
plot(fig1)    # point pattern
data(Helianthemum)
cosa12 <- K1K2(Helianthemum, j="deadpl", i="survpl", r=seq(0,200,le=201),
               nsim=99, nrank=1, correction="isotropic")
plot(cosa12$k1k2, lty=c(2, 1, 2), col=c(2, 1, 2), xlim=c(0, 200),
     main= "survival- death",ylab=expression(K[1]-K[2]), legend=FALSE)

```


Geostatistics
-------------

Useful packages:

```{r message=FALSE, eval=FALSE}
library(gstat)
library(geoR)
library(akima)   # for spline interpolation
# etc (see Spatial Task View)

library(spdep)   # dealing with spatial dependence

```



INTERACTING AND COMMUNICATING WITH OTHER GIS
============================================

```{r message=F, eval=F}
library(spgrass6)   # GRASS
library(RPyGeo)     # ArcGis (Python)
library(RSAGA)      # SAGA
library(spsextante) # Sextante 

```


Other useful packages
=====================

```{r message=FALSE, eval=FALSE}
library(Metadata)    # automatically collates data from online GIS datasets (land cover, pop density, etc) for a given set of coordinates

#library(GeoXp)    # Interactive exploratory spatial data analysis
example(columbus)
histomap(columbus,"CRIME")

library(maptools)
# readGPS 

library(rangeMapper)    # plotting species distributions, richness and traits


# Species Distribution Modelling
library(dismo)
library(BIOMOD)
library(biomod2)
library(SDMTools)

library(BioCalc)   # computes 19 bioclimatic variables from monthly climatic values (tmin, tmax, prec)

```




To learn more
=============

Packages help and vignettes, especially

http://cran.r-project.org/web/packages/raster/vignettes/Raster.pdf   
http://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf   
http://cran.r-project.org/web/packages/sp/vignettes/sp.pdf   

CRAN Task View: Analysis of Spatial Data
http://cran.r-project.org/web/views/Spatial.html

R-SIG-Geo mailing list
https://stat.ethz.ch/mailman/listinfo/R-SIG-Geo

R wiki: tips for spatial data
http://rwiki.sciviews.org/doku.php?id=tips:spatial-data&s=spatial

book
http://www.asdar-book.org/  

R spatial tips: http://spatialanalysis.co.uk/r/   
  
http://spatial-analyst.net   


